---
title: "CACTI: peak-based data preparation & mapping"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CACTI: peak-based data preparation & mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                     error = FALSE, cache = FALSE)
# Display each distinct warning once; leave all computations and errors unchanged.
warning_hook <- knitr::knit_hooks$get("warning")
shown_warnings <- character()
knitr::knit_hooks$set(warning = function(x, options) {
  x <- unique(x)
  x <- x[!x %in% shown_warnings]
  if (!length(x)) return("")
  shown_warnings <<- c(shown_warnings, x)
  warning_hook(x, options)
})
```

## Introduction

CACTI groups nearby peaks into windows and tests their joint association with genetic variants. This vignette describes the inputs, a one-function quick start, the output files, and the individual analysis steps.

Two input modes are available:

- **Genotype input:** provide genotype, phenotype, phenotype metadata, and covariates. CACTI runs single-peak mapping with MatrixEQTL before testing windows.
- **Summary-statistics input:** provide precomputed SNP–peak association statistics instead of genotype. Phenotypes and covariates are still required to estimate the correlation between peaks.

The main function is `cacti_peak_window()`. By default, `chr = "All"` processes all chromosomes present in the phenotype metadata, and `do_fdr = TRUE` adds window-level q-values.

## Part 1: Overview of the CACTI pipeline

### Workflow

1. **Map single-peak cis-QTLs.** Run MatrixEQTL on genotype, phenotype, and covariates. Skip this step when summary statistics are supplied.
2. **Group peaks into windows.** Assign peaks to non-overlapping genomic windows using their start positions.
3. **Adjust phenotypes for covariates.** Use the residualized phenotypes to estimate correlations between peaks.
4. **Test windows and add FDR.** Windows with two or more peaks use the principal-component omnibus test. Included single-peak windows use a univariate test. Aggregate the results and add window-level q-values.

| Function | Role |
|:--|:--|
| `cacti_peak_window()` | Runs the complete workflow. |
| `cacti_matrixqtl_cis()` | Generates single-peak cis-QTL summary statistics. |
| `cacti_group_peak_window()` | Groups peaks into windows. |
| `cacti_pheno_cov_residual()` | Adjusts phenotypes for covariates. |
| `cacti_cal_p()` | Tests SNP–window associations for one chromosome. |
| `cacti_add_fdr()` | Summarizes window results and adds q-values. |

### Input files

| Input | Format |
|:--|:--|
| `file_pheno_meta` | Four columns, in this order: `phe_id`, `phe_chr`, `phe_from`, `phe_to`. Use consistent chromosome labels, such as `chr5`. |
| `file_pheno` | Rows are peaks; the first column is the peak ID and the remaining columns are sample IDs. Supply normalized peak intensities. |
| `file_cov` | Rows are covariates; the first column is the covariate ID and the remaining columns are sample IDs. |
| `file_vcf` | Genotypes in VCF format. Alternatively, supply `file_geno` and `file_snp_pos`; see `?cacti_matrixqtl_cis`. |
| `qtl_file` or `qtl_files` | Alternative to genotype input: a table containing `phe_id`, `var_id`, and `z` (signed Z-scores). Include the associations to be tested, not only significant hits. |

Use matching peak and sample IDs across inputs. In summary-statistics mode, the association statistics, phenotypes, and covariates should describe the same analysis samples and features. For genome-wide analysis, provide the inputs for all chromosomes to be analyzed.

## Part 2: Quick start

### 1. Locate test data

The package includes a small chromosome 5 example. All inputs below are read from the installed package; results are written to a temporary directory.

```{r locate-peak-inputs}
library(cacti)

file_pheno_meta <- system.file("extdata", "test_cacti_peak_chr5_pheno_meta.bed", package = "cacti")
file_pheno <- system.file("extdata", "test_cacti_peak_chr5_pheno.txt", package = "cacti")
file_cov <- system.file("extdata", "test_cacti_peak_chr5_covariates.txt", package = "cacti")
file_vcf <- system.file("extdata", "test_cacti_peak_chr5_geno.vcf", package = "cacti")
qtl_file <- system.file("extdata", "test_cacti_peak_chr5_matrixqtl_sumstats.txt.gz", package = "cacti")

stopifnot(all(file.exists(c(file_pheno_meta, file_pheno, file_cov, file_vcf, qtl_file))))
print(basename(c(file_pheno_meta, file_pheno, file_cov, file_vcf, qtl_file)))

out_dir <- tempfile("cacti_peak_vignette_")
dir.create(out_dir)
out_prefix <- file.path(out_dir, "quickstart")
```

#### Phenotype metadata

```{r preview-peak-metadata}
knitr::kable(
  data.table::fread(file_pheno_meta, nrows = 5),
  caption = "Peak coordinates: first five rows."
)
```

#### Phenotype and covariate matrices

Only the first four samples are shown.

```{r preview-peak-matrices}
pheno_preview <- data.table::fread(file_pheno, nrows = 5)
cov_preview <- data.table::fread(file_cov, nrows = 5)

knitr::kable(
  pheno_preview[, 1:5, with = FALSE],
  digits = 3,
  caption = "Normalized peak intensities: first five peaks and four samples."
)
knitr::kable(
  cov_preview[, 1:5, with = FALSE],
  digits = 3,
  caption = "Covariates: first five rows and four samples."
)
```

### 2. Run the pipeline

Use 50-kb windows for peak grouping and a 100-kb cis distance for single-peak mapping. These are separate settings: `window_size` controls peak grouping, whereas `cis_dist` controls which SNP–peak pairs are tested by MatrixEQTL.

Identical warnings are displayed only once on this page. The example can produce an ACAT warning when a component P value equals 1; the final output P values are checked to be finite and within [0, 1].

```{r run-peak-genotype}
res <- cacti_peak_window(
  window_size = "50kb",
  file_pheno_meta = file_pheno_meta,
  file_pheno = file_pheno,
  file_cov = file_cov,
  file_vcf = file_vcf,
  cis_dist = 100000,
  chr = "All",
  min_peaks = 1,
  do_fdr = TRUE,
  out_prefix = out_prefix
)
```

In this example, `"All"` means chromosome 5 because only chromosome 5 is present in the input metadata.

### 3. Explore results

The function returns paths to the generated files.

```{r list-peak-outputs}
data.frame(
  output = names(res),
  file = vapply(res, function(x) paste(basename(x), collapse = "; "), character(1))
)
```

| Output | Contents |
|:--|:--|
| `file_peak_group` | Window coordinates and the peaks included in each window. |
| `file_peak_group_peaklevel` | Peak-to-window mapping. |
| `file_pheno_cov_residual` | Phenotype matrix after covariate adjustment. |
| `file_p_peak_group` | SNP–window association statistics, with columns `group`, `snp`, and `pval`. |
| `file_fdr_out` | Window-level summaries and q-values. |

All previews below are read from the files generated by the quick start.

#### Peak-to-window mapping

```{r preview-peak-groups}
knitr::kable(
  data.table::fread(res$file_peak_group_peaklevel, nrows = 5),
  caption = "Peak-to-window mapping: first five rows."
)
```

#### Covariate-adjusted phenotype matrix

```{r preview-peak-residuals}
residual_preview <- data.table::fread(res$file_pheno_cov_residual, nrows = 5)
knitr::kable(
  residual_preview[, 1:5, with = FALSE],
  digits = 3,
  caption = "Covariate-adjusted phenotypes: first five peaks and four samples."
)
```

#### SNP–window association statistics

```{r preview-peak-associations}
association_preview <- data.table::fread(res$file_p_peak_group[1], nrows = 5)
association_preview$pval <- format(association_preview$pval, digits = 3, scientific = TRUE)
knitr::kable(
  association_preview,
  caption = "SNP–window associations: first five rows, not ranked by significance."
)
```

#### Window-level FDR results

`cacti_add_fdr()` combines SNP–window P values within each window using ACAT and calculates q-values across the analyzed windows. The output also includes the minimum P value, its SNP ID, the number of tested SNPs, and a Bonferroni-adjusted minimum P value. The `q` column corresponds to `best_hit_acat`.

```{r preview-peak-fdr}
fdr_preview <- data.table::fread(res$file_fdr_out, nrows = 5)
for (column in c("best_hit", "best_hit_bonf", "best_hit_acat", "q")) {
  fdr_preview[[column]] <- format(fdr_preview[[column]], digits = 3, scientific = TRUE)
}
knitr::kable(
  fdr_preview,
  caption = "Window-level summaries: first five rows."
)
```

## Part 3: Step-by-step workflow

The calls below show the individual steps performed by the quick start. They are shown for users who want to run or modify one step separately. This vignette reuses the files generated above instead of rerunning the same analysis.

### Step 1: Map single-peak cis-QTLs

The genotype-input workflow writes this intermediate summary-statistics file:

```{r locate-generated-peak-statistics}
file_qtl_generated <- paste0(out_prefix, "_matrixqtl_cis_all_chrs.txt.gz")
stopifnot(file.exists(file_qtl_generated))
```

The corresponding single-peak mapping call is:

```{r peak-map-cis-recipe, eval=FALSE}
cacti_matrixqtl_cis(
  file_pheno = file_pheno,
  file_pheno_meta = file_pheno_meta,
  file_cov = file_cov,
  file_vcf = file_vcf,
  file_qtl_out = file_qtl_generated,
  cis_dist = 100000,
  p_threshold = 1.0
)
```

`p_threshold = 1.0` retains all tested associations. The output includes signed Z-scores (`z`) and P values (`pval`).

```{r preview-generated-peak-statistics}
generated_qtl_preview <- data.table::fread(file_qtl_generated, nrows = 5)
generated_qtl_preview$pval <- format(generated_qtl_preview$pval, digits = 3, scientific = TRUE)
knitr::kable(
  generated_qtl_preview,
  digits = 4,
  caption = "Generated single-peak summary statistics: first five rows."
)
```

### Step 2: Group peaks into windows

```{r peak-group-recipe, eval=FALSE}
cacti_group_peak_window(
  window_size = "50kb",
  file_pheno_meta = file_pheno_meta,
  file_peak_group = res$file_peak_group,
  file_peak_group_peaklevel = res$file_peak_group_peaklevel
)
```

The window table records the number of peaks in `n_pid_group` and their IDs in `pid_group`. The peak-level mapping is shown above.

### Step 3: Adjust phenotypes for covariates

```{r peak-residual-recipe, eval=FALSE}
cacti_pheno_cov_residual(
  file_pheno = file_pheno,
  file_cov = file_cov,
  file_pheno_cov_residual = res$file_pheno_cov_residual
)
```

CACTI uses the correlations between these residualized phenotypes in the multivariate test.

### Step 4: Test SNP–window associations

```{r peak-test-recipe, eval=FALSE}
cacti_cal_p(
  file_qtl_cis_norm = file_qtl_generated,
  chr = "chr5",
  file_peak_group = res$file_peak_group,
  file_pheno_cov_residual = res$file_pheno_cov_residual,
  file_p_peak_group = res$file_p_peak_group[1],
  min_peaks = 1
)
```

`min_peaks = 1` includes both single-peak and multi-peak windows. Set it to `2` to restrict testing to multi-peak windows. Within a multi-peak window, the test uses SNPs with association statistics available for every included peak.

### Step 5: Add window-level FDR

```{r peak-fdr-recipe, eval=FALSE}
cacti_add_fdr(
  file_all_pval = res$file_p_peak_group,
  file_fdr_out = res$file_fdr_out
)
```

For a genome-wide analysis, pass all chromosome-specific P-value files together so that q-values are calculated across the analyzed chromosomes.

## Part 4: Optional usage

### Use precomputed summary statistics

Provide `qtl_file` instead of `file_vcf`. This skips MatrixEQTL, but the phenotype and covariate inputs are still used for peak correlations. The following example runs this mode using the bundled summary-statistics file.

```{r preview-precomputed-peak-statistics}
qtl_preview <- data.table::fread(qtl_file, nrows = 5)
qtl_preview$pval <- format(qtl_preview$pval, digits = 3, scientific = TRUE)
knitr::kable(
  qtl_preview,
  digits = 4,
  caption = "Bundled single-peak summary statistics: first five rows."
)
```

```{r run-peak-summary}
res_summary <- cacti_peak_window(
  window_size = "50kb",
  file_pheno_meta = file_pheno_meta,
  file_pheno = file_pheno,
  file_cov = file_cov,
  qtl_file = qtl_file,
  chr = "All",
  min_peaks = 1,
  do_fdr = TRUE,
  out_prefix = file.path(out_dir, "summary_input")
)
```

```{r preview-summary-mode-results}
summary_fdr_preview <- data.table::fread(res_summary$file_fdr_out, nrows = 5)
for (column in c("best_hit", "best_hit_bonf", "best_hit_acat", "q")) {
  summary_fdr_preview[[column]] <- format(summary_fdr_preview[[column]], digits = 3, scientific = TRUE)
}
knitr::kable(
  summary_fdr_preview,
  caption = "Window-level results from summary-statistics input: first five rows."
)
```

The two modes return the same types of output files. These examples demonstrate workflow execution; they do not evaluate statistical power, calibration, or biological validity.

```{r check-peak-results, include=FALSE}
for (result in list(res, res_summary)) {
  values <- data.table::fread(result$file_p_peak_group[1])$pval
  qvalues <- data.table::fread(result$file_fdr_out)$q
  stopifnot(length(values) > 0, all(is.finite(values)),
            all(values >= 0 & values <= 1), length(qvalues) > 0,
            all(is.finite(qvalues)), all(qvalues >= 0 & qvalues <= 1))
}
```

### Select chromosomes or skip FDR

The following recipe uses `chr = "chr5"` for a targeted analysis and `do_fdr = FALSE` to return association P values without q-values. It is not rerun here because the quick-start inputs already contain only chromosome 5.

```{r peak-targeted-recipe, eval=FALSE}
res_chr <- cacti_peak_window(
  window_size = "50kb",
  file_pheno_meta = file_pheno_meta,
  file_pheno = file_pheno,
  file_cov = file_cov,
  qtl_file = qtl_file,
  chr = "chr5",
  do_fdr = FALSE,
  out_prefix = file.path(out_dir, "chr5_only")
)
```

For several chromosomes, supply a character vector to `chr`. `qtl_files` accepts a combined summary-statistics file, one file per selected chromosome, or a filename template containing `{chr}`. Use `?cacti_peak_window`, `?cacti_run_chr`, and `?cacti_run_genome` for the full argument descriptions.
