---
title: "CACTI peak-window pipeline"
author: "Lili Wang, Zining Qi and Xuanyao Liu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CACTI peak-window pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

This vignette explains:

1. The expected **input file formats**.  
2. Running CACTI.  
3. The **output files** produced by the pipeline.

The default mode is **genotype-first**:
- provide genotype + phenotype + covariates,
- CACTI runs MatrixEQTL first,
- then runs the standard CACTI peak-window testing.

Summary-statistics mode is optional.

---

## 1) Expected input file formats

### Required inputs (default mode)

1. **Genotype**
- `file_vcf`, or
- `file_geno` + `file_snp_pos`

2. **Phenotype matrix** (`file_pheno`)
- rows = peaks/features
- first column = feature ID (`ID`)
- remaining columns = sample IDs

3. **Phenotype metadata** (`file_pheno_meta`)
- required columns:
  - `phe_id`
  - `phe_chr`
  - `phe_from`
  - `phe_to`

4. **Covariate matrix** (`file_cov`)
- rows = covariates
- first column = covariate ID
- remaining columns = sample IDs

```{r show-required-inputs, message=FALSE}
library(cacti)

file_pheno_meta <- system.file("extdata", "test_pheno_meta.bed", package = "cacti")
file_pheno <- system.file("extdata", "test_pheno.txt", package = "cacti")
file_cov <- system.file("extdata", "test_covariates.txt", package = "cacti")
file_vcf <- system.file("extdata", "test_geno.vcf", package = "cacti")

head(data.table::fread(file_pheno_meta))
head(data.table::fread(file_pheno))[, 1:5]
head(data.table::fread(file_cov))[, 1:5]
```

### Optional input (summary-statistics mode)

If you already have cis-QTL summary stats, you can provide:
- `qtl_files` (or `qtl_file`) with required columns:
  - `phe_id`
  - `var_id`
  - `z`

```{r show-optional-qtl, message=FALSE}
qtl_file <- system.file("extdata", "test_qtl_sum_stats_chr5.txt.gz", package = "cacti")
head(data.table::fread(qtl_file))
```

---

## 2) Running CACTI

Use one function: `cacti_peak_window()`.

### A. Default mode (recommended): genotype input

This runs MatrixEQTL first, then standard CACTI automatically.

```{r run-cacti-default, eval=FALSE}
library(cacti)

res <- cacti_peak_window(
  window_size = "50kb",
  file_pheno_meta = file_pheno_meta,
  file_pheno = file_pheno,
  file_cov = file_cov,
  file_vcf = file_vcf,       # default genotype-first mode
  chrs = "chr5",
  cis_dist = 100000,         # 100 kb cis window for MatrixEQTL
  p_threshold = 1.0,         # keep all cis summary stats
  out_prefix = tempfile("cacti_peak_window_")
)

res
```

### B. Optional mode: summary stats already available

If you already have QTL summary statistics, pass `qtl_files` (or `qtl_file`).
In this mode, MatrixEQTL is skipped.

```{r run-cacti-summary-optional, eval=TRUE}
res_summary <- cacti_peak_window(
  window_size = "50kb",
  file_pheno_meta = file_pheno_meta,
  file_pheno = file_pheno,
  file_cov = file_cov,
  chrs = "chr5",
  qtl_file = qtl_file,       # optional summary-stat mode
  out_prefix = tempfile("cacti_peak_window_summary_")
)

res_summary
```

---

## 3) Output files produced by the pipeline

`cacti_peak_window()` returns the same output structure as `cacti_run_genome()`:

- `file_peak_group`
  - grouped windows table
- `file_peak_group_peaklevel`
  - peak-to-window mapping
- `file_pheno_cov_residual`
  - residualized phenotype matrix
- `file_p_peak_group`
  - per-chromosome `(group, snp, pval)` files
- `file_fdr_out`
  - final FDR-added window-level results

```{r show-outputs, eval=TRUE}
names(res_summary)
res_summary$file_fdr_out
```

For detailed low-level APIs, see:
- `?cacti_peak_window`
- `?cacti_run_chr`
- `?cacti_run_genome`
- `?cacti_matrixqtl_cis`
