---
title: "CACTI-S: segment-based data preparation & mapping"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CACTI-S: segment-based data preparation & mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  cache = FALSE
)
# Display each distinct warning once; leave all computations and errors unchanged.
warning_hook <- knitr::knit_hooks$get("warning")
shown_warnings <- character()
knitr::knit_hooks$set(warning = function(x, options) {
  x <- unique(x)
  x <- x[!x %in% shown_warnings]
  if (!length(x)) return("")
  shown_warnings <<- c(shown_warnings, x)
  warning_hook(x, options)
})
library(cacti)
```

## Introduction

The **CACTI-S (Segment-based)** pipeline is designed to process raw sequencing data (BAM files) into normalized phenotype matrices and QTL mapping. It handles the multi-step pipeline:

-   **Segmentation**: Tile the genome into fixed-size segments (e.g., 1kb, 5kb).
-   **Reads counting**: Count reads from raw BAM files into these segments.
-   **Preprocessing**: Perform quality control and normalization.
-   **Mapping**: Run cis-QTL mapping of the segments using `MatrixEQTL` to generate summary statistics required by the multivariate CACTI model.

This vignette explains:

1.  The overall workflow.
2.  The expected input files.
3.  The output files produced by the pipeline.
4.  Example usage with bundled toy data under `inst/extdata/`.

The main user-facing function in this vignette is:

-   `cacti_s_prep()` — a master wrapper that automates the entire workflow from BAM files to summary statistics.

Lower-level building blocks are:

| Function                  | Role        | Description                                                             |
|:-----------------------|:-----------------------|:-----------------------|
| **`cacti_s_prep`**        | **Wrapper** | Master function that runs the entire pipeline (Steps 1–4) sequentially. |
| `cacti_s_create_segments` | Step 1      | Generates fixed-size segments to divide the genome.                     |
| `cacti_s_count`           | Step 2      | Quantifies reads in segments using `Rsubread::featureCounts`.           |
| `cacti_s_preprocess`      | Step 3      | Performs QC and normalization.                                          |
| `cacti_s_map_cis`         | Step 4      | Runs cis-QTL mapping using `MatrixEQTL`.                                |

## Part 1: Overview of the CACTI-S pipeline

The CACTI-S pipeline has three main steps:

1.  **Define and quantify genomic segments**

-   Segmentation: Tiling the genome into fixed-size segments (e.g., 1kb, 5kb) or using a custom BED file.

-   Quantification: Counting reads in these segments for every sample (BAM file) using `Rsubread`.

-   Output: A raw counts matrix (Rows = Segments, Columns = Samples).

2.  **Preprocess and normalize phenotypes**

-   QC & Filtering & Normalization

-   Output: A normalized phenotype matrix ready for QTL mapping.

3.  **Cis-QTLs mapping**

-   Input: Normalized phenotype matrix, genotype data (VCF), and covariates.

-   Process: Runs MatrixEQTL for every variant–segment pair within a defined cis window (e.g., 100kb).

-   Output: Cis-QTL summary statistics (signed association statistics), which serve as the primary input for the CACTI multivariate test.

### Optional input-data QC

These independent functions report diagnostics and warnings without modifying inputs. They are not called automatically by the preprocessing or mapping workflow.

| Function | Input format | Diagnostics |
|:--|:--|:--|
| `cacti_qc_samples()` | Numeric feature-by-sample matrix, with `input_type = "raw_counts"` or `"normalized"`. | Missing/non-finite values; for raw counts, total counts for the sample and proportion of zero measurements. |
| `cacti_qc_features()` | Numeric feature-by-sample matrix, with the input type specified. | Missing/non-finite values and variance; for raw counts, mean counts, detection frequency, and proportion of zero measurements. |
| `cacti_qc_library()` | Named character vector of BAM paths; names are sample/library IDs. No index is required. | Read depth, mapping quality, usable reads, duplicate flags, mean base quality, and proportion of bases with Q >= 30. |

Matrix inputs are numeric R matrices or numeric data frames, not file paths. Samples are columns, with sample IDs as column names; features are rows, with feature IDs as row names. Both matrix functions return `n_samples`, `n_features`, a descriptive `summary` table, and `warnings`, without classifying features as pass/fail. Count totals, zero proportions, detection proportions, and variance use observed finite measurements; missing and non-finite values are reported separately. Metrics requiring observations are `NA` when none are available. Count totals sum over the supplied features, not all sequenced reads. Use `input_type = "normalized"` for normalized phenotypes; count-based diagnostics are then omitted. Warning thresholds are configurable: no features or samples are removed. These checks do not establish statistical power or false-positive control.

## Part 2: Quick start

For most users, the `cacti_s_prep()` function is the most convenient way to run the pipeline. It automates all steps and manages intermediate files.

### 1. Locate test data

We will use the small test files bundled with the `cacti` package. The four BAM files contain distinct count profiles across twelve 5-kb segments. These synthetic data demonstrate execution, not statistical power or biological validity. In a real analysis, replace these with your own BAM and VCF paths.

Identical warnings are displayed only once on this page. The generated association statistics and P values are checked to be finite.

```{r}
# Locate the 4 test BAM files
bam_files <- list.files(
  system.file("extdata", package = "cacti"),
  pattern = "Sample.*\\.bam$",
  full.names = TRUE
)

# Locate genotype and covariate files
file_vcf   <- system.file("extdata", "test_geno.vcf", package = "cacti")
file_cov   <- system.file("extdata", "test_cov.txt", package = "cacti")

print(basename(c(bam_files, file_vcf, file_cov)))
```

#### Optional: inspect the BAM files

Name each BAM path with its sample/library ID. Metrics count primary read records, not paired-end fragments. Usable reads are mapped, pass `min_mapq`, are not flagged as QC failures, and, by default, are not flagged as duplicates. The mapping-quality distribution is also returned.

```{r bam-input-qc}
named_bams <- setNames(bam_files, sub("\\.bam$", "", basename(bam_files)))
bam_qc <- cacti_qc_library(named_bams, min_mapq = 10)
knitr::kable(bam_qc$summary[, c(
  "sample_id", "total_reads", "mapped_reads", "usable_reads"
)])
knitr::kable(bam_qc$summary[, c(
  "sample_id", "mean_mapq", "mapq_pass_prop", "duplicate_prop"
)])
knitr::kable(bam_qc$summary[, c(
  "sample_id", "mean_base_quality", "prop_bases_q30",
  "reads_without_base_quality"
)])
```

Selected columns are shown here; `bam_qc$summary` contains the full set of diagnostics. `mapq_pass_prop` is the proportion of mapped reads passing `min_mapq`. MAPQ describes alignment confidence, not nucleotide base quality.

`mean_base_quality` is the mean Phred score across bases with stored quality scores, not a mean of per-read averages. `prop_bases_q30` is the proportion of these bases with Q >= 30. Base quality describes confidence in nucleotide calls. These metrics use all primary records, including unmapped reads and reads excluded by the usable-read filters, but not secondary or supplementary alignments. The report also includes `bases_with_quality` and `reads_without_base_quality`. Missing qualities are excluded, not treated as zero; the function warns when qualities are missing and returns `NA` for both quality metrics when no scores are available.

Duplicate-marking status is unknown by default, so the duplicate proportion is unavailable. Set `duplicate_flags = "marked"` only when duplicate marking is known to have been performed. Previously filtered BAMs cannot reveal the original read depth or the reads already removed, including duplicates. No library-complexity or signal-enrichment measures are calculated.

### 2. Run the pipeline

We will run the pipeline with the following settings:

-   **Genome**: hg19\
-   **Segment Size**: 5kb\
-   **Filter Mode**: `cpm_threshold` (keep segments with peak intensity CPM \> 1 in \> 20% of samples)

```{r}
out_dir <- file.path(tempdir(), "cacti_s_example")
res <- cacti_s_prep(
  file_bams = bam_files,
  file_vcf = file_vcf,
  file_cov = file_cov,
  
  out_dir = out_dir,
  out_prefix = "test",

  # Pipeline Parameters
  genome = "hg19",
  segment_size = "5kb",
  filter_mode = "cpm_threshold",
  min_cpm = 1,
  min_prop = 0.2,

  # Use 1 thread for this small demo
  threads = 1
)
```

### 3. Explore results

The function returns a list of paths to the generated files. Outputs are written to a temporary directory in this example; use a directory of your choice to keep them. The table below shows the filenames.

```{r}
knitr::kable(data.frame(Output = names(res), Filename = basename(unlist(res)), row.names = NULL))
```

#### Segments

This file contains the segment positions that divide the whole genome.

```{r}
saf <- read.table(res$saf, header = TRUE, nrows = 5)
knitr::kable(saf, caption = "Segment position (Top 5 Rows)")
```

#### Raw counts

This file contains raw counts within the segments.

```{r}
raw_counts <- data.table::fread(res$raw_counts, nrows = 5, data.table = FALSE)
names(raw_counts)[7:ncol(raw_counts)] <- basename(names(raw_counts)[7:ncol(raw_counts)])
knitr::kable(raw_counts, caption = "Raw counts within segments (Top 5 Rows)")
```

#### Optional: inspect the raw-count matrix

The following optional recipe reads all rows, removes the six annotation columns from the featureCounts table, and assigns segment and sample IDs to the numeric matrix. Unlike the five-row preview above, QC should use all supplied features. The library-size diagnostic is the total counts for the sample across these features, not the total number of sequenced reads.

```{r raw-count-qc, eval=FALSE}
count_table <- data.table::fread(res$raw_counts, data.table = FALSE)
count_matrix <- as.matrix(count_table[, 7:ncol(count_table), drop = FALSE])
rownames(count_matrix) <- count_table[[1]]
colnames(count_matrix) <- sub("\\.bam$", "", basename(colnames(count_matrix)))
sample_qc <- cacti_qc_samples(count_matrix, input_type = "raw_counts")
feature_qc <- cacti_qc_features(
  count_matrix, input_type = "raw_counts", min_count = 1, min_prop = 0.2
)
c(n_samples = sample_qc$n_samples, n_features = feature_qc$n_features)
sample_qc$summary
head(feature_qc$summary)
```

Here, detection means a raw count of at least 1. `min_prop = 0.2` triggers a warning when fewer than 20% of observed samples have detectable counts; it is not the CPM filter used by preprocessing and does not remove features. For real data, ensure the sample IDs extracted from the count-table column names match your genotype and covariate IDs.

#### Normalized phenotype matrix

This file contains filtered and normalized phenotypes, with segments in rows and samples in columns. Segment positions are stored in the separate metadata file.

```{r}
pheno <- read.table(res$pheno, header = TRUE)
knitr::kable(head(pheno, 5), digits = 3, caption = "First five normalized segment phenotypes")
```

#### Cis-QTL summary statistics

This file contains the cis-QTL mapping associations of individual segments, with signed Z-scores (`z`) and P values (`pval`).

```{r}
if (!is.null(res$qtl_stats)) {
  qtl <- data.table::fread(res$qtl_stats, data.table = FALSE)
  qtl_preview <- head(qtl, 5)
  qtl_preview$pval <- formatC(qtl_preview$pval, format = "e", digits = 3)
  knitr::kable(qtl_preview, digits = 3, caption = "First five SNP-segment associations")
}
```

```{r prep-summary}
knitr::kable(data.frame(
  Output = c("Retained segments", "SNP-segment associations"),
  Number = c(nrow(pheno), nrow(qtl))
))
```

```{r check-preparation, include=FALSE}
stopifnot(nrow(pheno) > 0, nrow(qtl) > 0,
          all(is.finite(as.matrix(pheno[-1]))), all(is.finite(qtl$z)),
          all(is.finite(qtl$pval)), all(qtl$pval >= 0 & qtl$pval <= 1),
          max(abs(2 * pnorm(-abs(qtl$z)) - qtl$pval)) < 1e-12)
```

## Part 3: Step-by-step workflow

If you need more control (e.g., inspecting intermediate files) or customization, you can run the individual functions included in the whole pipeline separately.

### Step 1: Create segments

First, we generate a SAF (Simplified Annotation Format) file that defines the genomic segments.

```{r}
file_saf <- file.path(out_dir, "test_segments.saf")

cacti_s_create_segments(
  file_saf_out = file_saf, # Output path for the SAF file.
  genome = "hg19", # Genome build to use: "hg19" (default) or "hg38".
  segment_size = "5kb" # Character like "5kb" or numeric bp (e.g. 5000).
)
```

```{r}
head(read.table(file_saf, header = TRUE))
```

### Step 2: Count Reads

Next, we quantify reads falling into these segments for all BAM files.

```{r}
file_counts <- file.path(out_dir, "test_raw_counts.txt")

cacti_s_count(
  file_bams = bam_files, # Character vector of BAM file paths.
  file_saf = file_saf, # Path to the SAF annotation file (from \code{cacti_s_create_segments}).
  file_counts_out = file_counts, # Output path for the counts matrix.
  threads = 1 # Number of threads for featureCounts.
)
```

```{r}
head(read.table(file_counts, header = TRUE))
```

### Step 3: Preprocess

We then perform preprocessing steps on the count data, including -

-   **QC:** Remove segments with 0 reads in all samples.\
-   **CPM:** Transform counts to log2(CPM + 0.5).\
-   **Filter:** Remove noise segments based on `filter_mode`. We use the default mode: `cpm_threshold` (keep segments with peak intensity CPM \> 1 in \> 20% of samples). We also provide other ways for filtering -
    -   `match_overlap_count`: Determines $N$, the number of segments overlapping regions in `filter_bed`. Then keeps the top $N$ segments by average intensity.
    -   `prop`: Keeps the top `prop_top` proportion of segments by average intensity.
-   **Standardize:** Z-score normalization per feature (Mean = 0, SD = 1).\
-   **RankNorm:** Inverse Normal Transform per sample.

```{r}
file_pheno <- file.path(out_dir, "test_pheno_norm.txt")
file_pheno_meta <- file.path(out_dir, "test_pheno_norm_meta.txt")

cacti_s_preprocess(
  file_raw_counts = file_counts, # Path to featureCounts output (from \code{cacti_s_count}).
  file_out = file_pheno, # Output path for the processed phenotype file.
  file_out_meta = file_pheno_meta, # Output path for the meta file of processed phenotype.
  filter_mode = "cpm_threshold", # Strategy to select segments: "cpm_threshold" (default).
  min_cpm = 1, # (For "cpm_threshold") Minimum CPM threshold (default 1).
  min_prop = 0.2 # (For "cpm_threshold") Minimum proportion of samples (default 0.2).
)
```

```{r}
head(read.table(file_pheno, header = TRUE))
```

```{r}
head(read.table(file_pheno_meta, header = TRUE))
```

### Step 4: Map Cis-QTLs

Finally, we run the cis-QTL mapping using the normalized phenotype matrix using `MatrixEQTL`.

```{r}
file_qtl <- file.path(out_dir, "test_cis_qtl_stats.txt")

cacti_s_map_cis(
  file_pheno = file_pheno, # Path to the processed phenotype file (from `cacti_s_preprocess`).
  file_pheno_meta = file_pheno_meta, # Path for the meta file of processed phenotype.
  file_cov = file_cov, # Path to covariate matrix.
  file_vcf = file_vcf, # Path to input VCF file.
  file_qtl_out = file_qtl, # Output path for summary statistics.
  cis_dist = 100000, # Cis-window distance (default 100000 bp = 100kb).
  p_threshold = 1.0 # P-value threshold for output (default 1.0, print all associations).
)
```

```{r}
head(read.table(file_qtl, header = TRUE))
```

## Part 4: Run multivariate association with CACTI

The files generated by the CACTI-S pipeline serve as the primary inputs for the core CACTI analysis. These include the phenotype metadata (`file_pheno_meta`), normalized phenotypes (`file_pheno`), covariates (`file_cov`), and the QTL summary statistics (`file_qtl`). For more details, please see the vignette [here](cacti_peak_window.html).

With these inputs ready, we can directly run `cacti_run_chr()` (for a single chromosome) or `cacti_run_genome()` (for the whole genome).

For the bundled example, 10-kb windows group adjacent 5-kb segments, allowing windows with at least two retained segments to enter the multivariate test.

```{r}
file_pheno <- file.path(out_dir, "test_pheno_norm.txt")
file_pheno_meta <- file.path(out_dir, "test_pheno_norm_meta.txt")
file_cov <- system.file("extdata", "test_cov.txt", package = "cacti")
file_qtl <- file.path(out_dir, "test_cis_qtl_stats.txt")
out_prefix <- file.path(out_dir, "cactis_chr1")
chr <- "chr1"

res_single_chr <- cacti_run_chr(
  window_size     = "10kb",
  file_pheno_meta = file_pheno_meta,
  file_pheno      = file_pheno,
  file_cov        = file_cov,
  chr             = chr,
  qtl_file        = file_qtl,
  out_prefix      = out_prefix,
  min_peaks       = 2
)
```

The function returns a list containing the paths to the generated files:

```{r}
knitr::kable(data.frame(Output = names(res_single_chr), Filename = basename(unlist(res_single_chr)), row.names = NULL))
```

Below is a preview of the final association statistics:

```{r}
window_results <- data.table::fread(res_single_chr$file_p_peak_group, data.table = FALSE)
window_preview <- head(window_results, 5)
window_preview$pval <- formatC(window_preview$pval, format = "e", digits = 3)
knitr::kable(window_preview, caption = "First five SNP-window associations")
knitr::kable(data.frame(
  Output = c("Tested windows", "SNP-window associations"),
  Number = c(length(unique(window_results$group)), nrow(window_results))
))
```

```{r check-windows, include=FALSE}
groups <- data.table::fread(res_single_chr$file_peak_group)
stopifnot(nrow(window_results) > 0, all(is.finite(window_results$pval)),
          all(window_results$pval > 0 & window_results$pval <= 1),
          all(groups$n_pid_group[match(window_results$group, groups$group)] >= 2))
```
